---
title: "Pb-210 Sediment Chronology — RERCA Summary Report"
subtitle: "Multi-model age-depth comparison and carbon accumulation readiness assessment"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    embed-resources: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)

# Load snapshot written by Script 05 — must be in the same directory as this .qmd
load("05_report_data.RData")

flag_colour <- function(x) {
  ifelse(x == "acceptable",
         '<span style="color:#1a9641;font-weight:bold;">&#10004; acceptable</span>',
         '<span style="color:#d7191c;font-weight:bold;">&#9888; HIGH — more data needed</span>')
}
```

## Executive Summary

This report summarises the multi-model Pb-210 sediment chronology produced by
the RERCA workflow (scripts 01–05). Four independent age-depth methods were
compared: **pb210 CRS**, **rplum**, **serac CRS**, **serac CFCS**, and
**Bayesian (Stan)**. The report evaluates model agreement, flags sections where
chronological uncertainty is too large for reliable carbon accumulation
reconstruction, and prepares the age-depth dataset for carbon accumulation rate
(RERCA) calculation once organic carbon or LOI measurements are available.

> **Carbon data status:** No LOI or %OC measurements have been provided yet.
> A CAR-ready table (all non-carbon terms pre-calculated) has been saved as
> `05_car_ready_table.csv`. Add a `loi_percent` or `oc_percent` column and
> multiply by the `car_factor` column to compute RERCA.

---

## Pb-210 Activity Profile

```{r activity-plot}
#| fig-cap: "Total and unsupported Pb-210 activity with depth. Dashed line = estimated background."
if (!is.null(activity_df)) {
  bg_line <- unique(activity_df$pb210_supported)[1]

  act_long <- activity_df %>%
    select(depth_cm, total = pb210_total_dpm_g, unsupported = pb210_unsupported) %>%
    pivot_longer(c(total, unsupported), names_to = "fraction", values_to = "activity") %>%
    left_join(
      activity_df %>% select(depth_cm,
                             total_err = pb210_error_dpm_g,
                             unsup_err = pb210_unsupported_err),
      by = "depth_cm"
    ) %>%
    mutate(
      err      = ifelse(fraction == "total", total_err, unsup_err),
      fraction = recode(fraction,
                        total       = "Total Pb-210",
                        unsupported = "Unsupported Pb-210")
    )

  surface_act   <- activity_df$pb210_total_dpm_g[which.min(activity_df$depth_cm)]
  dynamic_range <- round(surface_act / bg_line, 1)

  ggplot(act_long, aes(x = activity, y = depth_cm, colour = fraction)) +
    geom_point(size = 2.5) +
    geom_errorbar(aes(ymin = depth_cm, ymax = depth_cm,
                      xmin = activity - err, xmax = activity + err),
                  width = 0.3) +
    geom_vline(xintercept = bg_line, linetype = "dashed", colour = "grey40") +
    annotate("text", x = bg_line, y = max(activity_df$depth_cm) * 0.9,
             label = paste0("Background\n", round(bg_line, 3), " DPM/g"),
             hjust = -0.1, size = 3.5, colour = "grey40") +
    scale_y_reverse(name = "Depth (cm)") +
    scale_x_continuous(name = "Activity (DPM / g dry wt)") +
    scale_colour_manual(values = c("Total Pb-210"       = "#2c7bb6",
                                   "Unsupported Pb-210" = "#d7191c")) +
    labs(colour = NULL) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
} else {
  cat("Activity profile data not available.")
}
```

```{r activity-table}
if (!is.null(activity_df)) {
  surface_act   <- activity_df$pb210_total_dpm_g[which.min(activity_df$depth_cm)]
  bg_line       <- unique(activity_df$pb210_supported)[1]
  dynamic_range <- round(surface_act / bg_line, 1)

  dr_note <- dplyr::case_when(
    dynamic_range >= 5  ~ "Robust signal — strong model constraints expected.",
    dynamic_range >= 3  ~ "Moderate signal — models should be reliable but consider additional measurements.",
    TRUE                ~ "Weak signal — dating uncertainty will be high; additional Pb-210 measurements strongly recommended."
  )

  knitr::kable(data.frame(
    Metric = c("Measured sections", "Background (supported)", "Surface total activity",
               "Dynamic range (surface/background)", "Assessment"),
    Value  = c(nrow(activity_df),
               paste0(round(bg_line, 3), " DPM/g"),
               paste0(round(surface_act, 3), " DPM/g"),
               paste0(dynamic_range, "x"),
               dr_note)
  ), col.names = c("Metric", "Value"))
}
```

---

## Age-Depth Model Comparison

```{r age-depth-plot}
#| fig-cap: "All Pb-210 age-depth models overlaid. Shaded bands = 95% uncertainty intervals."
ggplot(all_ages, aes(x = age, y = depth_cm, colour = model, fill = model)) +
  geom_ribbon(aes(xmin = age_min_yr, xmax = age_max_yr),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_reverse(name = "Depth (cm)") +
  scale_x_continuous(name = "Age (years before coring)") +
  scale_colour_manual(values = palette_colours) +
  scale_fill_manual(values = palette_colours, guide = "none") +
  labs(subtitle = "Shaded bands = 95% uncertainty intervals", colour = "Model") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold"))
```

### Age comparison at 2-cm depth intervals

```{r age-comparison-table}
age_wide %>%
  select(-starts_with("model_age_")) %>%
  mutate(across(where(is.numeric), \(x) round(x, 1))) %>%
  knitr::kable(caption = "Interpolated ages (yr before coring) at common 2-cm depth intervals.")
```

### Inter-model agreement

```{r agreement-table}
age_wide %>%
  select(depth_cm, model_age_min, model_age_max, model_age_range) %>%
  mutate(across(where(is.numeric), \(x) round(x, 1)),
         agreement = ifelse(model_age_range < 10, "good", "poor")) %>%
  knitr::kable(
    col.names = c("Depth (cm)", "Min age (yr)", "Max age (yr)", "Spread (yr)", "Agreement"),
    caption = "Spread < 10 yr = good model agreement; > 10 yr warrants investigation."
  )
```

---

## Model Diagnostics

```{r diagnostics-table}
ci_summary %>%
  mutate(
    flag       = ifelse(mean_ci_pct > 50, "wide", "ok"),
    n_high_unc = paste0(n_high_unc, " / ", n_sections, " sections")
  ) %>%
  knitr::kable(
    col.names = c("Model", "Sections dated", "Datable depth (cm)",
                  "Oldest age (yr BP)", "Mean 95% CI (% of age)",
                  "Sections with CI > 50%", "CI flag"),
    caption = paste0("95% CI > 50% of estimated age flags sections where uncertainty",
                     " is too large for reliable RERCA.")
  )
```

```{r serac-cfcs}
if (!is.null(serac_stats) && "cfcs_r2" %in% names(serac_stats)) {
  cfcs_row <- serac_stats[!is.na(serac_stats$cfcs_r2), ]
  if (nrow(cfcs_row) > 0) {
    r2    <- cfcs_row$cfcs_r2[1]
    sar   <- abs(cfcs_row$cfcs_sar_mm_yr[1])
    sar_e <- abs(cfcs_row$cfcs_sar_err[1])
    note  <- if (r2 > 0.9) {
      "R2 > 0.9: consistent with constant sedimentation. CFCS appropriate."
    } else if (r2 > 0.8) {
      "R2 0.8-0.9: moderate fit. CRS likely more appropriate than CFCS."
    } else {
      "R2 < 0.8: poor fit. Variable rates — CRS or Bayesian model preferred."
    }
    knitr::kable(
      data.frame(
        Metric = c("CFCS SAR (mm/yr)", "CFCS R2", "Interpretation"),
        Value  = c(paste0(round(sar, 3), " +/- ", round(sar_e, 3)), round(r2, 3), note)
      ),
      col.names = c("Metric", "Value"),
      caption = "CFCS linearity check (serac)."
    )
  }
}
```

```{r bayesian-diagnostics}
if (!is.null(bayes_stats) && "n_divergent" %in% names(bayes_stats)) {
  divs     <- bayes_stats$n_divergent[1]
  div_note <- if (divs == 0) {
    "0 divergences — sampler converged well."
  } else if (divs <= 10) {
    paste0(divs, " divergences — minor; consider tightening priors.")
  } else {
    paste0(divs, " divergences — poor MCMC mixing. Treat results cautiously.")
  }
  knitr::kable(
    data.frame(
      Metric = c("Divergent transitions", "phi (supported) mean", "phi 95% CI", "Assessment"),
      Value  = c(divs,
                 paste0(round(bayes_stats$phi_mean_dpm_g[1], 3), " DPM/g"),
                 paste0(round(bayes_stats$phi_95_lo[1], 3), " - ",
                        round(bayes_stats$phi_95_hi[1], 3), " DPM/g"),
                 div_note)
    ),
    col.names = c("Metric", "Value"),
    caption = "Bayesian (Stan) MCMC diagnostics."
  )
}
```

---

## Cs-137 Validation

```{r cs137-section}
#| results: asis
if (!has_cs137) {
  cat(
    "> **Cs-137 data not present in this dataset.**\n>\n",
    "> Adding Cs-137 measurements is strongly recommended as an independent\n",
    "> time marker. The 1963 peak (peak atmospheric fallout) provides a fixed\n",
    "> date to validate or anchor the Pb-210 age-depth models.\n>\n",
    "> **What to do:** Include `cs137_dpm_g` and `cs137_error_dpm_g` columns\n",
    "> in your input CSV and re-run scripts 01-05.\n"
  )
}
```

```{r cs137-plot}
#| eval: !expr has_cs137
#| fig-cap: "Cs-137 activity profile. The 1963 peak provides an independent time marker."
if (has_cs137) {
  cs_df      <- activity_df[!is.na(activity_df$cs137_dpm_g), ]
  peak_depth <- cs_df$depth_cm[which.max(cs_df$cs137_dpm_g)]

  age_at_peak <- all_ages %>%
    group_by(model) %>%
    summarise(
      age_at_cs137 = approx(depth_cm, age, xout = peak_depth, rule = 1)$y,
      .groups = "drop"
    ) %>%
    filter(!is.na(age_at_cs137)) %>%
    mutate(
      year_AD     = as.integer(format(Sys.Date(), "%Y")) - age_at_cs137,
      expected_yr = 1963,
      offset_yr   = round(year_AD - expected_yr, 1)
    )

  p_cs <- ggplot(cs_df, aes(x = cs137_dpm_g, y = depth_cm)) +
    geom_point(size = 2.5, colour = "#756bb1") +
    geom_errorbar(aes(ymin = depth_cm, ymax = depth_cm,
                      xmin = cs137_dpm_g - cs137_error_dpm_g,
                      xmax = cs137_dpm_g + cs137_error_dpm_g),
                  width = 0.3, colour = "#756bb1") +
    geom_hline(yintercept = peak_depth, linetype = "dashed", colour = "grey40") +
    annotate("text", x = max(cs_df$cs137_dpm_g) * 0.7, y = peak_depth,
             label = paste0("Peak at ", round(peak_depth, 1), " cm"),
             vjust = -0.5, size = 3.5, colour = "grey40") +
    scale_y_reverse(name = "Depth (cm)") +
    scale_x_continuous(name = "Cs-137 activity (DPM / g dry wt)") +
    theme_bw(base_size = 13)
  print(p_cs)

  knitr::kable(age_at_peak,
    col.names = c("Model", "Age at Cs-137 peak (yr BP)",
                  "Modelled year (AD)", "Expected year (AD)", "Offset (yr)"),
    caption = paste0("Cs-137 peak at ", round(peak_depth, 1),
                     " cm. Offset < 5 yr = good validation; > 10 yr warrants investigation.")
  )
}
```

---

## Carbon Accumulation Rate (RERCA) Readiness

### What is RERCA?

Recent Rates of Carbon Accumulation (RERCA) are calculated as:

$$
\text{RERCA} \; (\text{g C m}^{-2} \text{yr}^{-1}) =
C_{\text{fraction}} \times \rho_b \; (\text{g cm}^{-3}) \times
\text{SAR} \; (\text{cm yr}^{-1}) \times 10{,}000
$$

where $C_{\text{fraction}}$ is the organic carbon proportion (from LOI or
elemental analysis), $\rho_b$ is dry bulk density, and SAR is from the
age-depth model.

### CAR-ready table (carbon data pending)

All terms except the carbon fraction have been pre-calculated.
Multiply `car_factor` by your LOI-derived or measured carbon fraction to get RERCA.

```{r car-table}
if (!is.null(car_ready)) {
  car_ready %>%
    select(-ci_width_yr) %>%
    mutate(
      age_yr_BP          = round(age_yr_BP, 1),
      age_95_lo          = round(age_95_lo, 1),
      age_95_hi          = round(age_95_hi, 1),
      sed_rate_cm_yr     = round(sed_rate_cm_yr, 4),
      bulk_density_g_cm3 = round(bulk_density_g_cm3, 3),
      car_factor         = round(car_factor, 2),
      ci_pct_age         = paste0(ci_pct_age, "%")
    ) %>%
    knitr::kable(
      col.names = c("Depth (cm)", "Age (yr BP)", "95% lo", "95% hi",
                    "SAR (cm/yr)", "Bulk density (g/cm3)",
                    "CAR factor", "CI (% of age)", "Uncertainty flag"),
      caption = paste0(
        "car_factor = bulk_density x SAR x 10000. ",
        "Multiply by C_fraction to get RERCA (g C/m2/yr). ",
        "Flagged rows have 95% CI > 50% of age.")
    )
} else {
  cat("> CAR-ready table could not be built — CRS ages missing.\n")
}
```

```{r car-flag-summary}
#| results: asis
if (!is.null(car_ready)) {
  n_high <- sum(car_ready$uncertainty_flag != "acceptable", na.rm = TRUE)
  n_ok   <- sum(car_ready$uncertainty_flag == "acceptable", na.rm = TRUE)
  cat(paste0("\n**", n_ok, " of ", nrow(car_ready),
             " sections** have acceptable uncertainty (95% CI < 50% of age)."))
  if (n_high > 0) {
    bad_depths <- car_ready$depth_cm[car_ready$uncertainty_flag != "acceptable"]
    cat(paste0(" **", n_high, " section(s) flagged** at depths: ",
               paste(round(bad_depths, 1), collapse = ", "),
               " cm — additional measurements needed before using for RERCA.\n"))
  }
}
```

---

## What More Is Needed

Items ordered by priority.

### Required — not yet available

| Priority | Measurement | Why needed | Column in template |
|---|---|---|---|
| **1** | LOI (%) or %OC (elemental) | The carbon fraction in the RERCA equation | `loi_percent` / add `oc_percent` |
| **2** | Carbon for *all* dated sections | Gaps create unresolvable bias in the reconstruction | same |

### Strongly recommended — to reduce chronological uncertainty

| Priority | Measurement | Why needed | Column in template |
|---|---|---|---|
| **3** | Additional Pb-210 at high-uncertainty depths | Reduces CI width in flagged sections | `pb210_total_dpm_g` |
| **4** | Cs-137 activity profile | Independent 1963 time marker to validate the Pb-210 model | `cs137_dpm_g` |
| **5** | Ra-226 at multiple depths | Direct measurement of supported activity; removes tail-estimation uncertainty | `ra226_supported_dpm_g` |

### Optional — to strengthen the reconstruction

| Priority | Measurement | Why needed |
|---|---|---|
| **6** | Spheroidal carbonaceous particles (SCPs) | Additional stratigraphic markers (~1850, ~1950, ~1980) |
| **7** | Replicate Pb-210 measurements | Quantify analytical reproducibility near the detection limit |
| **8** | Core top Pb-210 if not measured | Anchors the model at the sediment-water interface |

---

## Next Steps

1. **Select a preferred age-depth model** from the comparison plot and diagnostics.
   Use CRS as the default for variable-sedimentation sites; CFCS only if R² > 0.9.

2. **Collect organic carbon data** (LOI% or %OC) for all sections in the
   CAR-ready table. If using LOI, apply a site-appropriate conversion factor
   (e.g. Craft et al. 1991 for coastal wetlands; Heiri et al. 2001 for lakes).

3. **Address flagged sections** — add Pb-210, Cs-137, or Ra-226 measurements
   at depths where uncertainty exceeds 50% of the estimated age.

4. **Add Cs-137 data** and re-run scripts 01–05. The validation section above
   will auto-populate once `cs137_dpm_g` is present in the input CSV.

5. **Calculate RERCA** — load `05_car_ready_table.csv`, join your carbon data,
   and compute: `RERCA = C_fraction * car_factor`.

6. **Propagate uncertainty** — carry the 95% CI on age through to the final
   RERCA estimates, especially for sections flagged as high-uncertainty.

---

## Session Info

```{r session-info}
sessionInfo()
```
